---
title: "Data Visualization"
author: "Chinazom"
date: "2024-04-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(modeldata)

```


```{r}
?crickets
View(crickets)
```

# The basics
## Creating a chart for temp and rate
### geom_point known as **scatter plot** is good for two quantitative variable
```{r}

ggplot(crickets, aes(x = temp, 
                     y = rate)) + 
  geom_point()
```
 
## To label the chart
```{r}
ggplot(crickets, aes(x = temp, 
                     y = rate)) + 
  geom_point() +
  labs(x = "Temperature",
       y = "Chirp rate",
       title = "Cricket chirps",
       caption = "Source: McDonald (2009)")
```

#### To change the color of the charts based on the different species, you add color = species

```{r}
ggplot(crickets, aes(x = temp, 
                     y = rate,
                     color = species)) + 
  geom_point() +
  labs(x = "Temperature",
       y = "Chirp rate",
       color = "Species",
       title = "Cricket chirps",
       caption = "Source: McDonald (2009)")
```

### To get a darker shade of the color using scale_color_brewer
```{r}
ggplot(crickets, aes(x = temp, 
                     y = rate,
                     color = species)) + 
  geom_point() +
  labs(x = "Temperature",
       y = "Chirp rate",
       color = "Species",
       title = "Cricket chirps",
       caption = "Source: McDonald (2009)") +
       scale_color_brewer(palette = "Dark2")
```

### Modifiying basic properties of the plot
```{r}
ggplot(crickets, aes(x = temp, 
                     y = rate)) + 
  geom_point(color = "blue",
             size = 2,
             shape = "square") +
  labs(x = "Temperature",
       y = "Chirp rate",
       title = "Cricket chirps",
```


```{r}
caption = "Source: McDonald (2009)")

```


## Adding another layer geom_smooth to show the direction of the relationship
```{r}

ggplot(crickets, aes(x = temp, 
                     y = rate)) + 
  geom_point() +
  geom_smooth() +
  labs(x = "Temperature",
       y = "Chirp rate",
       title = "Cricket chirps",
       caption = "Source: McDonald (2009)")

```

### To modify the geom_smooth
```{r}
ggplot(crickets, aes(x = temp, 
                     y = rate)) + 
  geom_point() +
  geom_smooth(method = "lm", # to make the line linear
              se = FALSE) + # se = false takes away the shadow from the line
  labs(x = "Temperature",
       y = "Chirp rate",
       title = "Cricket chirps",
       caption = "Source: McDonald (2009)")
```
```{r}
ggplot(crickets, aes(x = temp, 
                     y = rate,
                     color = species)) + # to differentiate between the different species 
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(x = "Temperature",
       y = "Chirp rate",
       color = "Species",
       title = "Cricket chirps",
       caption = "Source: McDonald (2009)") +
  scale_color_brewer(palette = "Dark2") # to make the colors darker
```
## Other plots
### Histogram- mostly use for one quantitative variable 

```{r}
ggplot(crickets, aes(x = rate)) + 
  geom_histogram(bins = 15) #  bin represents a group or a segment of the data, interval into which the range of data values is divided
```


### Geom_freqpoly -  uses lines to represent the frequency distribution of a continuous variable.

```{r}
ggplot(crickets, aes(x = temp)) + 
  geom_freqpoly(bins = 15)
```
### Bar chart mostly used for one categorical variable

```{r}
ggplot(crickets, aes(x = species)) + 
  geom_bar()
```
### Modification of the Bar chart output
#### To change the outer and inner color of the bar
```{r}
ggplot(crickets, aes(x = species)) + 
  geom_bar (color = "black", # color function works for the borders of the chart
fill = "lightblue") # fill for the inside
```
### To give the values in a variable different colors - here species

```{r}
ggplot(crickets, aes(x = species, 
                     fill = species)) + 
  geom_bar(show.legend = FALSE) + # to remove the legend from the chart 
  scale_fill_brewer(palette = "Dark2") # to make the fill darker 
```


```{r}
ggplot(crickets, aes(x = species, y = count)) + 
  geom_bar(stat = "identity")

```

### Geom_boxplot is good for one categorical and one quantitative variables
#### Boxplot is a visual representation of your 5 number summary; minimum - the line below
#### 1st quartile, the 2nd quartile which is the medium (the line in-between ), the 3rd quartile and the max - the line above

```{r}
ggplot(crickets, aes(x = species, 
                     y = rate,
                     color = species)) + 
  geom_boxplot(show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2") # To make the color darker
```

### Theme() - control all non-data display. Use theme() if you just need to tweak the display of an existing, we have  theme.minimal,classic,bw,light etc

#### To remove the gray background in the chart using theme minimal
```{r}
ggplot(crickets, aes(x = species, 
                     y = rate,
                     color = species)) + 
  geom_boxplot(show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

### Faceting

#### Without Faceting
```{r}
ggplot(crickets, aes(x = rate, 
                     fill = species)) + 
  geom_histogram(bins = 15) +
  scale_fill_brewer(palette = "Dark2")
```

#### With faceting
```{r}

ggplot(crickets, aes(x = rate,
                     fill = species)) + 
  geom_histogram(bins = 15,
                 show.legend = FALSE) + 
  facet_wrap(~species) +
  scale_fill_brewer(palette = "Dark2")
```
#### To have just one column - the second bar will be below the first 
```{r}

ggplot(crickets, aes(x = rate,
                     fill = species)) + 
  geom_histogram(bins = 15,
                 show.legend = FALSE) + 
  facet_wrap(~species,
             ncol = 1) +
  scale_fill_brewer(palette = "Dark2") + 
  theme_minimal()
```


