---
title: "R Class"
output: html_notebook
---

# Introduction
Transforming a  wide data it into 5 columns: Store, Category, Date, Sales, and Margin, by unpivoting and pivoting, while handling the dates, metric, and sales/margin columns.

### Step 1: Load the necessary libraries

```{r}
library(tidyverse)
```


### Step 2: Read the dataset

```{r}
# using the import dataset gui
p <- read.csv("~/Excel datsets/PQ Exercise 1d.csv")
   print(p)
```


```{r}
#coping as path, you either use double backward slash \\ or a single forward slash

 read.csv("C:\\Users\\ADMIN\\Documents\\Excel datsets\\PQ Exercise 1d.csv", stringsAsFactors = FALSE)
```


### Step 3: Unpivot the dataset using pivot_longer()
Unpivoting will convert the date columns into rows under a single Date column. Before doing that, let's rename the first few columns for clarity.


```{r}
# Rename the first columns to proper names
colnames(p)[1:3] <- c("Metric", "Store", "Category")

# Pivot longer, converting date columns into a single column 'Date'
 P <- p %>%
  pivot_longer(cols = contains("X"), names_to = "Date", values_to = "Value") 

```


```{r}
# removing the quater columns
p <- P %>% 
  select(-starts_with("Q"))
```


### Step 4: Convert Date into a Date format
First we need to convert the date strings from the format X01.01.2017 to 1/1/2017 by removing prefix "x" and replacing periods "." with slash using the "gsub()" function


```{r}
p$Date <- gsub("^X", "", p$Date)  # Remove the leading 'X'
p$Date <- gsub("\\.", "/", p$Date)  # Replace '.' with '/'

```




```{r}
# Convert 'Date' column to Date format (the dates are in 'm/d/yyyy' format)
 p$Date <- as.Date(p$Date, format = "%m/%d/%Y")

```


### Step 5: Remove non-date values in the Date column
Remove rows where the Date column contains non-date values like "Q1 2017".


```{r}
# Filter out rows with non-date values (like 'Q1 2017')
any(is.na(p$Date))
p <- p %>% filter(!is.na(Date))

```


## Step 6: Filling up blanks in column 'metric'

```{r}
# Convert blanks to NA
p <- p %>%
  mutate(Metric = na_if(Metric, ""))  # Convert empty strings to NA

# Fill down the 'Metric' column to handle NA values
p <- p %>%
  fill(Metric, .direction = "down")

```


### Step 7: Pivoting metric and values to create different columns for sales and margin

```{r}
# Spread the 'Metric' column into 'Sales' and 'Margin'
p <- p %>%
  pivot_wider(names_from = Metric, values_from = Value)
```


```{r}
# Filter out rows where both 'Sales' and 'Margin' are NA
p <- p %>%
  filter(!(is.na(Sales) & is.na(Margin)))

view(p)

```


## Analysis and Visualization


```{r}
# Summarize total sales and average margin by Store
 p %>%
  group_by(Store) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Avg_Margin = mean(Margin, na.rm = TRUE)
  )

```


```{r}
# Summarize total sales and average margin by Category 
 p %>%
  group_by(Category) %>%
  summarise(
    Total_Sales = sum(Sales),
    Avg_Margin = mean(Margin)
  )
```


```{r}
```


```{r}
```


```{r}
# Visualize how sales vary over time.
ggplot(p, aes(x = Date, y = Sales, color = as.factor(Store))) +
  geom_line() +
  labs(title = "Sales Over Time",
       x = "Date",
       y = "Sales",
       color = "Store") +
  theme_minimal()

```


```{r}
# Total Sales by Store
ggplot(p, aes(x = as.factor(Store), y = Sales, fill = as.factor(Store))) +
  geom_bar(stat = "identity") +
  labs(title = "Total Sales by Store",
       x = "Store",
       y = "Total Sales") +
  theme_minimal()

# Average Margin by Store
ggplot(p, aes(x = as.factor(Store), y = Avg_Margin, fill = as.factor(Store))) +
  geom_bar(stat = "identity") +
  labs(title = "Average Margin by Store",
       x = "Store",
       y = "Average Margin") +
  theme_minimal()

```























#### Summarize the Data by Month and Year

```{r}
# Summarize total sales and average margin by Month and Year
summary_by_month_year <- p %>%
  group_by(Year, Month) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Avg_Margin = mean(Margin, na.rm = TRUE)
  ) %>%
  arrange(Year, Month)

```


```{r}
```


```{r}
```


```{r}
```

